---
title: "NBA Draft, Players & Performance - Analysis - R Notebook"
output: html_notebook
---

# Analyzing the NBA Drafts, Players & First-Year Performance 1990-2020 dataset

## Objectives

1.  Calculate z-scores for pure scoring performance (average points per game) for number one picks of each draft.
2.  Conduct t-tests for the following:
    -   Comparing the average height of players drafted in the first round versus those drafted in the second round.
    -   Comparing the average net rating of players across different draft rounds
3.  Conduct ANOVA + post hoc tests for the following:
    -   Comparing the average height of players across different continents
    -   Comparing the net rating of players across different continents
4.  Conduct chi-square tests for the following:
    -   Association between draft number and country
    -   Association between draft number and position
    -   Association between draft number and college

<!-- -->

5.  Calculate the correlations for the following:
    -   Player height and net rating
    -   Draft number and net rating
6.  Build a predictive model that takes in the most significant numerical variables to predict performance (net rating?).

## Load the data

```{r}
setwd("/Users/thomaslee/Desktop/Data/Projects/NBA Draft, Players & Performance/data")
nba_data <- read.csv("NBA_cleaned_data.csv")
```

## View the data

```{r}
nba_data
```

## View the structure of the data

```{r}
str(nba_data)
```

## View the head and tail of the data

```{r}
head(nba_data)
tail(nba_data)
```

## Calculate z-scores for first-year scoring performance of number one draft picks from 1990-2020

```{r}
# Calculate the mean for points (pts) scored by freshly drafted players for all draft years.
mean_1990 <- mean(nba_data[nba_data$draft_year == 1990, 'pts'])
mean_1991 <- mean(nba_data[nba_data$draft_year == 1991, 'pts'])
mean_1992 <- mean(nba_data[nba_data$draft_year == 1992, 'pts'])
mean_1993 <- mean(nba_data[nba_data$draft_year == 1993, 'pts'])
mean_1994 <- mean(nba_data[nba_data$draft_year == 1994, 'pts'])
mean_1995 <- mean(nba_data[nba_data$draft_year == 1995, 'pts'])
mean_1996 <- mean(nba_data[nba_data$draft_year == 1996, 'pts'])
mean_1997 <- mean(nba_data[nba_data$draft_year == 1997, 'pts'])
mean_1998 <- mean(nba_data[nba_data$draft_year == 1998, 'pts'])
mean_1999 <- mean(nba_data[nba_data$draft_year == 1999, 'pts'])
mean_2000 <- mean(nba_data[nba_data$draft_year == 2000, 'pts'])
mean_2001 <- mean(nba_data[nba_data$draft_year == 2001, 'pts'])
mean_2002 <- mean(nba_data[nba_data$draft_year == 2002, 'pts'])
mean_2003 <- mean(nba_data[nba_data$draft_year == 2003, 'pts'])
mean_2004 <- mean(nba_data[nba_data$draft_year == 2004, 'pts'])
mean_2005 <- mean(nba_data[nba_data$draft_year == 2005, 'pts'])
mean_2006 <- mean(nba_data[nba_data$draft_year == 2006, 'pts'])
mean_2007 <- mean(nba_data[nba_data$draft_year == 2007, 'pts'])
mean_2008 <- mean(nba_data[nba_data$draft_year == 2008, 'pts'])
mean_2009 <- mean(nba_data[nba_data$draft_year == 2009, 'pts'])
mean_2010 <- mean(nba_data[nba_data$draft_year == 2010, 'pts'])
mean_2011 <- mean(nba_data[nba_data$draft_year == 2011, 'pts'])
mean_2012 <- mean(nba_data[nba_data$draft_year == 2012, 'pts'])
mean_2013 <- mean(nba_data[nba_data$draft_year == 2013, 'pts'])
mean_2014 <- mean(nba_data[nba_data$draft_year == 2014, 'pts'])
mean_2015 <- mean(nba_data[nba_data$draft_year == 2015, 'pts'])
mean_2016 <- mean(nba_data[nba_data$draft_year == 2016, 'pts'])
mean_2017 <- mean(nba_data[nba_data$draft_year == 2017, 'pts'])
mean_2018 <- mean(nba_data[nba_data$draft_year == 2018, 'pts'])
mean_2019 <- mean(nba_data[nba_data$draft_year == 2019, 'pts'])
mean_2020 <- mean(nba_data[nba_data$draft_year == 2020, 'pts'])

# Calculate the standard deviation for points (pts) scored by freshly drafted players for all draft years.
sd_1990 <- sd(nba_data[nba_data$draft_year == 1990, 'pts'])
sd_1991 <- sd(nba_data[nba_data$draft_year == 1991, 'pts'])
sd_1992 <- sd(nba_data[nba_data$draft_year == 1992, 'pts'])
sd_1993 <- sd(nba_data[nba_data$draft_year == 1993, 'pts'])
sd_1994 <- sd(nba_data[nba_data$draft_year == 1994, 'pts'])
sd_1995 <- sd(nba_data[nba_data$draft_year == 1995, 'pts'])
sd_1996 <- sd(nba_data[nba_data$draft_year == 1996, 'pts'])
sd_1997 <- sd(nba_data[nba_data$draft_year == 1997, 'pts'])
sd_1998 <- sd(nba_data[nba_data$draft_year == 1998, 'pts'])
sd_1999 <- sd(nba_data[nba_data$draft_year == 1999, 'pts'])
sd_2000 <- sd(nba_data[nba_data$draft_year == 2000, 'pts'])
sd_2001 <- sd(nba_data[nba_data$draft_year == 2001, 'pts'])
sd_2002 <- sd(nba_data[nba_data$draft_year == 2002, 'pts'])
sd_2003 <- sd(nba_data[nba_data$draft_year == 2003, 'pts'])
sd_2004 <- sd(nba_data[nba_data$draft_year == 2004, 'pts'])
sd_2005 <- sd(nba_data[nba_data$draft_year == 2005, 'pts'])
sd_2006 <- sd(nba_data[nba_data$draft_year == 2006, 'pts'])
sd_2007 <- sd(nba_data[nba_data$draft_year == 2007, 'pts'])
sd_2008 <- sd(nba_data[nba_data$draft_year == 2008, 'pts'])
sd_2009 <- sd(nba_data[nba_data$draft_year == 2009, 'pts'])
sd_2010 <- sd(nba_data[nba_data$draft_year == 2010, 'pts'])
sd_2011 <- sd(nba_data[nba_data$draft_year == 2011, 'pts'])
sd_2012 <- sd(nba_data[nba_data$draft_year == 2012, 'pts'])
sd_2013 <- sd(nba_data[nba_data$draft_year == 2013, 'pts'])
sd_2014 <- sd(nba_data[nba_data$draft_year == 2014, 'pts'])
sd_2015 <- sd(nba_data[nba_data$draft_year == 2015, 'pts'])
sd_2016 <- sd(nba_data[nba_data$draft_year == 2016, 'pts'])
sd_2017 <- sd(nba_data[nba_data$draft_year == 2017, 'pts'])
sd_2018 <- sd(nba_data[nba_data$draft_year == 2018, 'pts'])
sd_2019 <- sd(nba_data[nba_data$draft_year == 2019, 'pts'])
sd_2020 <- sd(nba_data[nba_data$draft_year == 2020, 'pts'])

# Calculate the z-scores for number one picks for all draft years.
d_coleman <- (18.1 - mean_1990) / sd_1990
l_johnson <- (12.8 - mean_1991) / sd_1991
s_neal <- (26.2 - mean_1992) / sd_1992
c_webber <- (20.1 - mean_1993) / sd_1993
g_robinson <- (21.1 - mean_1994) / sd_1994
j_smith <- (18.7 - mean_1995) / sd_1995
a_iverson <- (23.5 - mean_1996) / sd_1996
t_duncan <- (21.1 - mean_1997) / sd_1997
m_olowokandi <- (8.9 - mean_1998) / sd_1998
e_brand <- (20.1 - mean_1999) / sd_1999
k_mart <- (12 - mean_2000) / sd_2000
k_brown <- (4.5 - mean_2001) / sd_2001
y_ming <- (13.5 - mean_2002) / sd_2002
l_james <- (20.9 - mean_2003) / sd_2003
d_howard <- (12 - mean_2004) / sd_2004
a_bogut <- (9.4 - mean_2005) / sd_2005
a_bargnani <- (11.6 - mean_2006) / sd_2006
g_oden <- (8.9 - mean_2007) / sd_2007
d_rose <- (16.8 - mean_2008) / sd_2008
b_griffin <- (22.5 - mean_2009) / sd_2009
j_wall <- (16.4 - mean_2010) / sd_2010
k_irving <- (18.5 - mean_2011) / sd_2011
a_davis <- (13.5 - mean_2012) / sd_2012
a_bennett <- (4.2 - mean_2013) / sd_2013
a_wiggins <- (16.9 - mean_2014) / sd_2014
k_towns <- (18.3 - mean_2015) / sd_2015
b_simmons <- (15.8 - mean_2016) / sd_2016
m_fultz <- (7.1 - mean_2017) / sd_2017
d_ayton <- (16.3 - mean_2018) / sd_2018
z_williamson <- (22.5 - mean_2019) / sd_2019
a_edwards <- (19.3 - mean_2020) / sd_2020

# Create initial unsorted dataframe with player names and z-scores.
df <- data.frame(
  Player_Name = c("Derrick Coleman", "Larry Johnson", "Shaquille O'Neal", "Chris Webber", "Glenn Robinson", "Joe Smith", "Allen Iverson", "Tim Duncan", "Michael Olowokandi", "Elton Brand", "Kenyon Martin", "Kwame Brown", "Yao Ming", "LeBron James", "Dwight Howard", "Andrew Bogut", "Andrea Bargnani", "Greg Oden", "Derrick Rose", "Blake Griffin", "John Wall", "Kyrie Irving", "Anthony Davis", "Anthony Bennett", "Andrew Wiggins", "Karl-Anthony Towns", "Ben Simmons", "Markelle Fultz", "Deandre Ayton", "Zion Williamson", "Anthony Edwards"),
  draft_year = c(1990:2020),
  points_scored = c(18.1, 12.8, 26.2, 20.1, 21.1, 18.7, 23.5, 21.1, 8.9, 20.1, 12, 4.5, 13.5, 20.9, 12, 9.4, 11.6, 8.9, 16.8, 22.5, 16.4, 18.5, 13.5, 4.2, 16.9, 18.3, 15.8, 7.1, 16.3, 22.5, 19.3),
  z_scores = c(d_coleman, l_johnson, s_neal, c_webber, g_robinson, j_smith, a_iverson, t_duncan, m_olowokandi, e_brand, k_mart, k_brown, y_ming, l_james, d_howard, a_bogut, a_bargnani, g_oden, d_rose, b_griffin, j_wall, k_irving, a_davis, a_bennett, a_wiggins, k_towns, b_simmons, m_fultz, d_ayton, z_williamson, a_edwards),
  stringasAsFactors = FALSE
)

# Sort initial dataframe and output final dataframe sorted by z-scores descending.
df <- df[order(df$z_scores, decreasing = TRUE),]
print(df)
```

Important note: the z-scores reflect each player's scoring performance for their first year among other players who were also drafted that year, NOT all players.

Some surprising findings here. Although generally players with top average scored points are among those with top z-scores, the rankings are quite different between the two.

What's particularly interesting is that Ben Simmons and John Wall have #2 and #3 z-scores respectively, although their average scored points are sub-20. Whereas Shaquille has the highest average scored points but ranks unexpectedly at #17 for z-scores. This calls for closer inspection.

It may be that the players in Shaquille's draft class were simply more competitive: four players averaged 20+ points and 17 players averaged at least double digits. Whereas in 2016 (the year Ben Simmons was drafted): zero freshly drafted players averaged 20+ points and only three averaged double digits.

According to these findings, the z-scores may actually be more of a reflection of each player's draft class talent pool and competitiveness than just the player's individual talent.

## Conduct t-test to compare the average height of players drafted in the first round versus those drafted in the second round

```{r}
# Install and load "dplyr" package
library(dplyr)

# Make sure 'draft_round' is a factor or character variable for filtering
nba_data$draft_round <- as.character(nba_data$draft_round)

# Filter data for players in the first and second rounds
first_round_heights <- nba_data %>%
  filter(draft_round == "1") %>%
  select(player_height) %>%

second_round_heights <- nba_data %>%
  filter(draft_round == "2") %>%
  select(player_height) %>%

# Conduct the independent samples t-test
t_test_result <- t.test(first_round_heights$player_height, second_round_heights$player_height, var.equal = FALSE)

# Print the results of the t-test
print(t_test_result)
```

The p-value of 0.01192 indicates that if there were no real difference in the average heights of first and second-round players, the probability of observing the resulting difference in sample means due to random chance would be about 1.19%. This low probability provides evidence against the null hypothesis.

In short, **there is statistically significant evidence to suggest that there is a true difference in the mean player heights between players selected in the first round and players selected in the second round** in the data.

To put it in less technical language, height does seem to be a factor, though far from the only one, in a player's success in the draft (in terms of being selected earlier).

## Conduct t-test to compare the average net rating of players across different draft rounds

```{r}
# Convert 'draft_round' to character for filtering
nba_data$draft_round <- as.character(nba_data$draft_round)

# Filter data for players in the first and second rounds, excluding Undrafted
first_round_net_ratings <- nba_data %>%
  filter(draft_round == "1") %>%
  select(net_rating)

second_round_net_ratings <- nba_data %>%
  filter(draft_round == "2") %>%
  select(net_rating)

# Conduct the independent samples t-test
t_test_result <- t.test(first_round_net_ratings$net_rating, second_round_net_ratings$net_rating, var.equal = FALSE)

# Print the results of the t-test
print(t_test_result)
```

The p-value of 0.001203 strongly indicates that if there were no real difference in the average net rating of first and second-round players, the probability of observing the resulting difference in sample means due to random chance would be about 0.12%. This low probability provides evidence against the null hypothesis.

In short, **there is statistically significant evidence to suggest that there is a true difference in the mean net ratings between players selected in the first round and players selected in the second round** in the data.

In simpler terms, the "promise" of players selected in the first round tends to be fulfilled since they generally perform better than players selected in the second round. This finding is intuitive, but it's good to get confirmation that team management and scouting departments are doing their jobs.

That said, there are always surprise performers; second round picks and even undrafted players can become all stars.

## Conduct ANOVA + post hoc tests to compare the average height of drafted players across different continents

```{r}
# install.packages("agricolae") # for post hoc test
# install.packages("countrycode") # to convert country to continent

library(agricolae)
library(countrycode)

# Convert 'draft_round' to character for filtering
nba_data$draft_round <- as.character(nba_data$draft_round)

# Create a 'continent' column based on the 'country' column
nba_data$continent <- countrycode(sourcevar = nba_data$country,
                                   origin = "country.name",
                                   destination = "continent")

# Filter data for drafted players, and remove missing values
df_filtered_rounds_continent <- nba_data %>%
  filter(draft_round %in% c("1", "2")) %>%  # Filter for first and second rounds
  filter(!is.na(player_height) & !is.na(continent)) # Remove missing player_height or continent

# Make sure 'continent' is a factor variable
df_filtered_rounds_continent$continent <- as.factor(df_filtered_rounds_continent$continent)

# Conduct the ANOVA test
anova_result_continent_filtered <- aov(player_height ~ continent, data = df_filtered_rounds_continent)

# Print the summary of the ANOVA results
summary(anova_result_continent_filtered)

# If the ANOVA result is significant (p < 0.05), conduct a post hoc test
if (summary(anova_result_continent_filtered)[[1]]$`Pr(>F)`[1] < 0.05) {
  posthoc_result_continent_filtered <- HSD.test(anova_result_continent_filtered, "continent")
  print(posthoc_result_continent_filtered)
} else {
  print("ANOVA result is not significant. No post hoc test is needed.")
}
```

The ANOVA test comparing the average player height across different continents yielded a statistically significant result. This indicates that **there is a significant difference in the average player height among at least one of the continents compared**. The large F-value and the extremely small p-value provide strong evidence against the null hypothesis that the average player height is the same across all continents.

According to the results of the Tukey HSD post hoc test, there is a statistically significant difference in the average height of NBA players across the different continents. **Specifically, players from Asia, Africa, and Europe tend to be significantly taller than players from the Americas**. Players from Oceania have an average height that falls in between and is not significantly different from the taller groups (Asia, Africa, Europe) or the shorter group (Americas) at the 0.05 significance level.

## Conduct ANOVA + post hoc tests to compare the average net rating of drafted players across different continents

```{r}
# install.packages("agricolae") # for post hoc test
# install.packages("countrycode") # to convert country to continent

library(agricolae)
library(countrycode)

# Convert 'draft_round' to character for filtering
nba_data$draft_round <- as.character(nba_data$draft_round)

# Create a 'continent' column based on the 'country' column
nba_data$continent <- countrycode(sourcevar = nba_data$country,
                                   origin = "country.name",
                                   destination = "continent")

# Filter data for drafted players, and remove missing values
df_filtered_rounds_continent <- nba_data %>%
  filter(draft_round %in% c("1", "2")) %>%  # Filter for first and second rounds
  filter(!is.na(net_rating) & !is.na(continent)) # Remove missing net_rating or continent

# Make sure 'continent' is a factor variable
df_filtered_rounds_continent$continent <- as.factor(df_filtered_rounds_continent$continent)

# Conduct the ANOVA test
anova_result_continent_filtered <- aov(net_rating ~ continent, data = df_filtered_rounds_continent)

# Print the summary of the ANOVA results
summary(anova_result_continent_filtered)

# If the ANOVA result is significant (p < 0.05), conduct a post hoc test
if (summary(anova_result_continent_filtered)[[1]]$`Pr(>F)`[1] < 0.05) {
  posthoc_result_continent_filtered <- HSD.test(anova_result_continent_filtered, "continent")
  print(posthoc_result_continent_filtered)
} else {
  print("ANOVA result is not significant. No post hoc test is needed.")
}
```

The results of the ANOVA test indicate that there is **no statistically significant difference** in the average net rating of players across the different continents. The observed variations in average net rating between the continental groups are not large enough to be considered statistically significant and could reasonably be attributed to random chance.

Note that these findings don't say anything about which continent produces the most drafted players; from a cursory glance of the dataset, it is clear that the USA produces the most NBA players by far. What these findings DO indicate is that among drafted players, their continent is not a significant factor behind their first-year performance.
