---
title: "NBA Draft, Players & Performance - Analysis - R Notebook"
output: html_notebook
---

# Analyzing the NBA Drafts, Players & Performance dataset

## Objectives

1.  Calculate z-scores for pure scoring performance (average points per game) for number one picks of each draft.
2.  Conduct t-tests for the following:
    -   Comparing the average height of players drafted in the first round versus those drafted in the second round.

    -   Comparing the average height of players from two different countries.

<!-- -->

3.  Conduct ANOVA + post hoc tests for the following:
    -   Average height of players across different draft rounds

    -   Comparing the average net rating of players across different draft rounds

    -   Comparing the average height or net rating of players from different continents or groups of countries

<!-- -->

4.  Conduct chi-square tests for the following:
    -   Association between draft number and country

    -   Association between draft number and position

    -   Association between draft number and college

<!-- -->

5.  Calculate the correlations for the following:
    -   Player height and net rating

    -   Draft number and net rating

<!-- -->

6.  Build a predictive model that takes in the most significant numerical variables to predict performance (net rating?).

## Load the data

```{r}
setwd("/Users/thomaslee/Desktop/Data/Projects/NBA Draft, Players & Performance/data")
nba_data <- read.csv("NBA_cleaned_data.csv")
```

## View the data

```{r}
nba_data
```

## View the structure of the data

```{r}
str(nba_data)
```

## View the head and tail of the data

```{r}
head(nba_data)
tail(nba_data)
```

## Calculate z-scores for first-year pure scoring performance of number one draft picks from 1990-2020
