---
title: "NBA Draft, Players & Performance - Analysis - R Notebook"
output: html_notebook
---

# Analyzing the NBA Drafts, Players & Performance dataset

## Objectives

1.  Calculate z-scores for pure scoring performance (average points per game) for number one picks of each draft.
2.  Conduct t-test for the following:
    -   Comparing the average height of players drafted in the first round versus those drafted in the second round.
3.  Conduct ANOVA + post hoc tests for the following:
    -   Comparing the average net rating of players across different draft rounds
    -   Comparing the average height or net rating of players from different continents or groups of countries
4.  Conduct chi-square tests for the following:
    -   Association between draft number and country
    -   Association between draft number and position
    -   Association between draft number and college

<!-- -->

5.  Calculate the correlations for the following:
    -   Player height and net rating
    -   Draft number and net rating
6.  Build a predictive model that takes in the most significant numerical variables to predict performance (net rating?).

## Load the data

```{r}
setwd("/Users/thomaslee/Desktop/Data/Projects/NBA Draft, Players & Performance/data")
nba_data <- read.csv("NBA_cleaned_data.csv")
```

## View the data

```{r}
nba_data
```

## View the structure of the data

```{r}
str(nba_data)
```

## View the head and tail of the data

```{r}
head(nba_data)
tail(nba_data)
```

## Calculate z-scores for first-year pure scoring performance of number one draft picks from 1990-2020

```{r}
# Calculate the mean for points (pts) scored by freshly drafted players for all draft years.
mean_1990 <- mean(nba_data[nba_data$draft_year == 1990, 'pts'])
mean_1991 <- mean(nba_data[nba_data$draft_year == 1991, 'pts'])
mean_1992 <- mean(nba_data[nba_data$draft_year == 1992, 'pts'])
mean_1993 <- mean(nba_data[nba_data$draft_year == 1993, 'pts'])
mean_1994 <- mean(nba_data[nba_data$draft_year == 1994, 'pts'])
mean_1995 <- mean(nba_data[nba_data$draft_year == 1995, 'pts'])
mean_1996 <- mean(nba_data[nba_data$draft_year == 1996, 'pts'])
mean_1997 <- mean(nba_data[nba_data$draft_year == 1997, 'pts'])
mean_1998 <- mean(nba_data[nba_data$draft_year == 1998, 'pts'])
mean_1999 <- mean(nba_data[nba_data$draft_year == 1999, 'pts'])
mean_2000 <- mean(nba_data[nba_data$draft_year == 2000, 'pts'])
mean_2001 <- mean(nba_data[nba_data$draft_year == 2001, 'pts'])
mean_2002 <- mean(nba_data[nba_data$draft_year == 2002, 'pts'])
mean_2003 <- mean(nba_data[nba_data$draft_year == 2003, 'pts'])
mean_2004 <- mean(nba_data[nba_data$draft_year == 2004, 'pts'])
mean_2005 <- mean(nba_data[nba_data$draft_year == 2005, 'pts'])
mean_2006 <- mean(nba_data[nba_data$draft_year == 2006, 'pts'])
mean_2007 <- mean(nba_data[nba_data$draft_year == 2007, 'pts'])
mean_2008 <- mean(nba_data[nba_data$draft_year == 2008, 'pts'])
mean_2009 <- mean(nba_data[nba_data$draft_year == 2009, 'pts'])
mean_2010 <- mean(nba_data[nba_data$draft_year == 2010, 'pts'])
mean_2011 <- mean(nba_data[nba_data$draft_year == 2011, 'pts'])
mean_2012 <- mean(nba_data[nba_data$draft_year == 2012, 'pts'])
mean_2013 <- mean(nba_data[nba_data$draft_year == 2013, 'pts'])
mean_2014 <- mean(nba_data[nba_data$draft_year == 2014, 'pts'])
mean_2015 <- mean(nba_data[nba_data$draft_year == 2015, 'pts'])
mean_2016 <- mean(nba_data[nba_data$draft_year == 2016, 'pts'])
mean_2017 <- mean(nba_data[nba_data$draft_year == 2017, 'pts'])
mean_2018 <- mean(nba_data[nba_data$draft_year == 2018, 'pts'])
mean_2019 <- mean(nba_data[nba_data$draft_year == 2019, 'pts'])
mean_2020 <- mean(nba_data[nba_data$draft_year == 2020, 'pts'])

# Calculate the standard deviation for points (pts) scored by freshly drafted players for all draft years.
sd_1990 <- sd(nba_data[nba_data$draft_year == 1990, 'pts'])
sd_1991 <- sd(nba_data[nba_data$draft_year == 1991, 'pts'])
sd_1992 <- sd(nba_data[nba_data$draft_year == 1992, 'pts'])
sd_1993 <- sd(nba_data[nba_data$draft_year == 1993, 'pts'])
sd_1994 <- sd(nba_data[nba_data$draft_year == 1994, 'pts'])
sd_1995 <- sd(nba_data[nba_data$draft_year == 1995, 'pts'])
sd_1996 <- sd(nba_data[nba_data$draft_year == 1996, 'pts'])
sd_1997 <- sd(nba_data[nba_data$draft_year == 1997, 'pts'])
sd_1998 <- sd(nba_data[nba_data$draft_year == 1998, 'pts'])
sd_1999 <- sd(nba_data[nba_data$draft_year == 1999, 'pts'])
sd_2000 <- sd(nba_data[nba_data$draft_year == 2000, 'pts'])
sd_2001 <- sd(nba_data[nba_data$draft_year == 2001, 'pts'])
sd_2002 <- sd(nba_data[nba_data$draft_year == 2002, 'pts'])
sd_2003 <- sd(nba_data[nba_data$draft_year == 2003, 'pts'])
sd_2004 <- sd(nba_data[nba_data$draft_year == 2004, 'pts'])
sd_2005 <- sd(nba_data[nba_data$draft_year == 2005, 'pts'])
sd_2006 <- sd(nba_data[nba_data$draft_year == 2006, 'pts'])
sd_2007 <- sd(nba_data[nba_data$draft_year == 2007, 'pts'])
sd_2008 <- sd(nba_data[nba_data$draft_year == 2008, 'pts'])
sd_2009 <- sd(nba_data[nba_data$draft_year == 2009, 'pts'])
sd_2010 <- sd(nba_data[nba_data$draft_year == 2010, 'pts'])
sd_2011 <- sd(nba_data[nba_data$draft_year == 2011, 'pts'])
sd_2012 <- sd(nba_data[nba_data$draft_year == 2012, 'pts'])
sd_2013 <- sd(nba_data[nba_data$draft_year == 2013, 'pts'])
sd_2014 <- sd(nba_data[nba_data$draft_year == 2014, 'pts'])
sd_2015 <- sd(nba_data[nba_data$draft_year == 2015, 'pts'])
sd_2016 <- sd(nba_data[nba_data$draft_year == 2016, 'pts'])
sd_2017 <- sd(nba_data[nba_data$draft_year == 2017, 'pts'])
sd_2018 <- sd(nba_data[nba_data$draft_year == 2018, 'pts'])
sd_2019 <- sd(nba_data[nba_data$draft_year == 2019, 'pts'])
sd_2020 <- sd(nba_data[nba_data$draft_year == 2020, 'pts'])

# Calculate the z-scores for number one picks for all draft years.
d_coleman <- (18.1 - mean_1990) / sd_1990
l_johnson <- (12.8 - mean_1991) / sd_1991
s_neal <- (26.2 - mean_1992) / sd_1992
c_webber <- (20.1 - mean_1993) / sd_1993
g_robinson <- (21.1 - mean_1994) / sd_1994
j_smith <- (18.7 - mean_1995) / sd_1995
a_iverson <- (23.5 - mean_1996) / sd_1996
t_duncan <- (21.1 - mean_1997) / sd_1997
m_olowokandi <- (8.9 - mean_1998) / sd_1998
e_brand <- (20.1 - mean_1999) / sd_1999
k_mart <- (12 - mean_2000) / sd_2000
k_brown <- (4.5 - mean_2001) / sd_2001
y_ming <- (13.5 - mean_2002) / sd_2002
l_james <- (20.9 - mean_2003) / sd_2003
d_howard <- (12 - mean_2004) / sd_2004
a_bogut <- (9.4 - mean_2005) / sd_2005
a_bargnani <- (11.6 - mean_2006) / sd_2006
g_oden <- (8.9 - mean_2007) / sd_2007
d_rose <- (16.8 - mean_2008) / sd_2008
b_griffin <- (22.5 - mean_2009) / sd_2009
j_wall <- (16.4 - mean_2010) / sd_2010
k_irving <- (18.5 - mean_2011) / sd_2011
a_davis <- (13.5 - mean_2012) / sd_2012
a_bennett <- (4.2 - mean_2013) / sd_2013
a_wiggins <- (16.9 - mean_2014) / sd_2014
k_towns <- (18.3 - mean_2015) / sd_2015
b_simmons <- (15.8 - mean_2016) / sd_2016
m_fultz <- (7.1 - mean_2017) / sd_2017
d_ayton <- (16.3 - mean_2018) / sd_2018
z_williamson <- (22.5 - mean_2019) / sd_2019
a_edwards <- (19.3 - mean_2020) / sd_2020

# Create initial unsorted dataframe with player names and z-scores.
df <- data.frame(
  Player_Name = c("Derrick Coleman", "Larry Johnson", "Shaquille O'Neal", "Chris Webber", "Glenn Robinson", "Joe Smith", "Allen Iverson", "Tim Duncan", "Michael Olowokandi", "Elton Brand", "Kenyon Martin", "Kwame Brown", "Yao Ming", "LeBron James", "Dwight Howard", "Andrew Bogut", "Andrea Bargnani", "Greg Oden", "Derrick Rose", "Blake Griffin", "John Wall", "Kyrie Irving", "Anthony Davis", "Anthony Bennett", "Andrew Wiggins", "Karl-Anthony Towns", "Ben Simmons", "Markelle Fultz", "Deandre Ayton", "Zion Williamson", "Anthony Edwards"),
  draft_year = c(1990:2020),
  points_scored = c(18.1, 12.8, 26.2, 20.1, 21.1, 18.7, 23.5, 21.1, 8.9, 20.1, 12, 4.5, 13.5, 20.9, 12, 9.4, 11.6, 8.9, 16.8, 22.5, 16.4, 18.5, 13.5, 4.2, 16.9, 18.3, 15.8, 7.1, 16.3, 22.5, 19.3),
  z_scores = c(d_coleman, l_johnson, s_neal, c_webber, g_robinson, j_smith, a_iverson, t_duncan, m_olowokandi, e_brand, k_mart, k_brown, y_ming, l_james, d_howard, a_bogut, a_bargnani, g_oden, d_rose, b_griffin, j_wall, k_irving, a_davis, a_bennett, a_wiggins, k_towns, b_simmons, m_fultz, d_ayton, z_williamson, a_edwards),
  stringasAsFactors = FALSE
)

# Sort initial dataframe and output final dataframe sorted by z-scores descending.
df <- df[order(df$z_scores, decreasing = TRUE),]
print(df)
```

Important note: the z-scores reflect each player's scoring performance for their first year among **other players who were also drafted that year**, NOT all players.

Some surprising findings here. Although generally players with top average scored points are among those with top z-scores, the rankings are quite different between the two.

What's particularly interesting is that Ben Simmons and John Wall have #2 and #3 z-scores respectively, although their average scored points are sub-20. Whereas Shaquille has the highest average scored points but ranks unexpectedly at #17 for z-scores. This calls for closer inspection.

It may be that the players in Shaquille's draft class were simply more competitive their first year: four players averaged 20+ points and 17 players averaged at least double digits. Whereas in 2016 (the year Ben Simmons was drafted): zero players averaged 20+ points and only three players averaged double digits.

According to these findings, the z-scores may actually be more of a reflection of each player's draft class talent pool and competitiveness than just the player's individual talent.
